### 2017-04-23 Manuscripts/cog-psych/cogpsych.tex
edits to creative cognition, future directions sections

\nbcnote{Make sure to write this!}

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{e2-yranges.pgf}
    \caption{Y-Axis range and position of the categories generated in Experiment 2. Each line corresponds to a single category, with notches corresponding to the Y-axis position of exemplars within the category (notches may overlap). Data from each participant is sorted by overall range, and then by condition. }
    \label{fig:e2-yranges}
    \end{center}
\end{figure}

@@

The creative generation of novel concepts is an intriguing yet understudied topic in cognitive science. While the bulk of prior research on the topic has focused on the classic finding that generated concepts tend to be distributionally similar to known concepts, there has been little work addressing the role of contrast in creative generation: how is it that people are able to create something {\em different} from what is already known? This issue is of fundamental importance: in order to successfully generate something new, it must be different from what is known. We developed a novel, exemplar-based model, PACKER, which formally specifies the role of contrast in generation. The model proposes that categories are represented as exemplars in a multidimensional psychological space, and generation is constrained both by within-category and between-category similarity: exemplars belonging to the same category should be similar to one another, and exemplars belonging to different categories should not be similar to one another.

In addition to an analysis of published data \citep[][Experiment 3]{jern2013probabilistic}, we reported two experiments demonstrating systematic effects of category contrast in creative generation: members of participant-generated categories tended to be highly dissimilar from members of previously-learned categories, and were usually more similar to one another than to members of other categories. Likewise, we observed broad interdependence between the distributional structure (feature variance, correlation) and physical instantiation (location within the domain space) of generated categories: In Experiment 2, we found that the unoccupied regions of the domain influenced the distributional structure of categories, and in both Experiments we observed that participants generated categories that were oriented to increase contrast with what was already known.

Finally, we conducted simulations comparing PACKER's account of our results to that of a ``copy-and-tweak'' model (realized as a variant of PACKER with no sensitivity to category contrast), and a hierarchical sampling model designed to explain the classic distributional similarity effect. In all simulations, we found that PACKER's sensitivity to contrast considerably enhanced the models account of generation. Further, by measuring PACKER's fit as a function of its prioritization of within- and between-category similarity, we observed that considering either constraint exclusively results in a relatively low-quality account. Instead, PACKER's best results were obtained when both constraints are considered, indicating that human learners do not generate novel concepts exclusively on the basis of within-category similarity or between class-contrast. This finding mirrors our behavioral results: it is clear that both constraints influence creative generation.

@@

The proposal that category contrast is a primary constraint in creative generation is, in some ways, entirely commonsense. In order to successful create something \textit{new}, it must be different from what is known. Beyond its role in creative generation, category contrast is also of fundamental importance in categorization more broadly: All other factors held constant, new categories are easier to learn if they are dissimilar to members of other categories, and knowledge of highly distinct categories is applied more accurately than that of ill-defined categories\footnote{\nbcnote{This is an uncontroversial claim but we should have some cites. I just can't think of a super appropriate one right now...}}. Likewise, the act of forming category representations affects similarity judgments about category members and nonmembers, with category members being viewed as more similar to one another than members of other categories \citep{goldstone1994influences,goldstone2001altering} .

@@

Indeed, issues relating to the trade-off between within- and between-class similarity arise in a variety of domains. For example, \cite{regier2007} showed that the partitioning of color categories reflects such a trade-off in a psychological space -- colors are partitioned into groups with members that are viewed as highly similar to one another yet distinct from other colors. A similar trade-off can be observed in phoneme perception... \nbcnote {I got nothing there...}

The above passages serve to demonstrate that the dual forces of within-class similarity and between-class contrast influence cognitive function in a wide variety of domains. The PACKER model is unique in that it extends this trade-off to the domain of creative generation, and allows us to begin to understand the processes involved in creative generation using our well-developed knowledge of human categorization.

@@

A central focus of the creative cognition approach \citep{finke1992creative,smith1995creative} has been to explain acts of creativity in terms of the mental processes that are commonly studied in cognitive psychology and cognitive science. The PACKER model advances this goal by applying the well-understood representational and process assumptions of the exemplar view to the act of generating new categories. In doing so, the model offers an additional interpretation of many of the central findings within the literature on creative cognition. For example, perhaps the most foundational principle from this literature concerns the limiting influence of prior knowledge: individuals create new categories composed of features from existing classes, and what is created can be influenced drastically through the introduction of cues or examples \citep{marsh1999inadvertent,smith1993constraining}. In this paper, we have identified another important aspect of the constraining influence of prior knowledge: what is generated cannot be the same as what is already known. Our results demonstrate that this influence is concisely explained in terms of a trade-off between within-category similarity and between-category dissimilarity.

Interestingly, the leading account within the creative cognition literature, the Path of Least Resistance account \citep{ward1994structured,ward1995s}, also explains generation in terms of an exemplar-based retrieval process. This account was designed to explain the creative generation of natural categories (e.g., new species of plants and animals) and as a result relies strongly on the hierarchical organization of these categories: individuals are thought to retrieve an example of the higher-level category being generated (e.g., \textit{bird} may be retrieved from the category \textit{animal}), and then systematically alter what was retrieved to make something new. As the PACKER model does not assume knowledge is hierarchically organized \citep[this is true of the exemplar view more boradly, see][]{murphy2016exemplar}, the model may be viewed as a formal instantiation of the Path of Least Resistance for application in a traditional artificial categorization domain (in which there is no established hierarchy of categories). PACKER's success in explaining generation within an artificial domain motivates future work exploring the nature of category contrast within a more naturalistic setting.

@@

Although an overall goal of this work is to advance the study of creative thought by explaining generation through quantitative approaches, PACKER is not easily aligned with many of the myriad alternative theoretical viewpoints in the study of creativity \citep[see][]{kozbelt2010theories}. Instead, in this paper we have studied a highly complex behavior (creative generation) as it applies within a well-established domain (category learning).

>>>>

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{e2-yranges.pgf}
    \caption{Y-Axis range and position of the participant-generated categories from Experiment 2. Each line corresponds to a participant's category, with notches corresponding to the Y-axis position of exemplars within the category (notches may overlap). Participants are sorted by overall range, and then by condition. }
    \label{fig:e2-yranges}
    \end{center}
\end{figure}

@@

The creative generation of novel concepts is an intriguing yet understudied topic in cognitive science. While the bulk of prior research on the topic has focused on the classic finding that generated concepts tend to be distributionally similar to known concepts, there has been little work addressing the role of contrast in creative generation: how is it that people are able to create something {\em different} from what is already known? This issue is of fundamental importance: In order to successfully generate something new, it must be in some sense distinct. We developed a novel, exemplar-based model, PACKER, which formally specifies the role of contrast in generation. The model proposes that categories are represented as exemplars in a multidimensional psychological space, and generation is constrained both by within-category and between-category similarity: exemplars belonging to the same category should be similar to one another, and exemplars belonging to different categories should not be similar to one another.

In addition to an analysis of published data \citep[][Experiment 3]{jern2013probabilistic}, we reported two experiments demonstrating systematic effects of category contrast in creative generation. Members of participant-generated categories tended to be highly dissimilar from members of previously-learned categories, and were usually more similar to one another than to members of other categories. We also observed broad interdependence between the distributional structure (feature variance, correlation) and physical instantiation (location within the domain space) of generated categories: In Experiment 2, we found that the unoccupied regions of the domain influenced the distributional structure of categories, and in both Experiments we observed that participants distributed their generated categories to increase contrast with what was already known.

We conducted simulations comparing PACKER's account of our results to that of a ``copy-and-tweak'' model (realized as a variant of PACKER with no sensitivity to category contrast), and a hierarchical sampling model designed to explain the classic distributional similarity effect. In all simulations, we found that PACKER's sensitivity to contrast considerably enhanced the models account of generation. Further, by measuring PACKER's fit as a function of its prioritization of within- and between-category similarity, we observed that considering either constraint exclusively results in a relatively low-quality account. Instead, PACKER's best results were obtained when both constraints are considered, indicating that human learners do not generate novel concepts exclusively on the basis of within-category similarity or between class-contrast. This finding mirrors our behavioral results and demonstrates that both constraints influence creative generation.

@@

The proposal that category contrast is a primary constraint in creative generation is, in some ways, entirely commonsense. In order to successful create something \textit{new}, it must be different from what is known. Beyond its role in creative generation, category contrast is also of fundamental importance in categorization more broadly. All other factors held constant, new categories are easier to learn if they are dissimilar to members of other categories, and knowledge of highly distinct categories is applied more accurately than that of ill-defined categories\footnote{\nbcnote{This is an uncontroversial claim but we should have some cites. I just can't think of a super appropriate one right now...}}. Likewise, the act of forming category representations affects similarity judgments about category members and nonmembers, with category members being viewed as more similar to one another than members of other categories \citep{goldstone1994influences,goldstone2001altering} .

@@

Even beyond the field of categorization, one can find instances of the trade-off between within and between-class similarity scattered throughout the research in cognitive science. For example, \cite{regier2007} showed that the partitioning of color categories reflects such a trade-off in a psychological space -- colors are partitioned into groups with members that are viewed as highly similar to one another yet distinct from other colors. A similar trade-off can be observed in phoneme perception... \nbcnote {I got nothing there...}

The above passages serve to demonstrate that the dual forces of within-class similarity and between-class contrast influence cognitive function in a wide variety of domains. The PACKER model is notable in that it successfully interprets this trade-off within the domain of creative generation, and allows us to begin to understand the relatively understudied processes involved in creative generation through our more well-developed knowledge of human categorization.

@@

Although the focus of this paper has been to address the processes involved in creative generation using the empirical and quantitative toolkit of traditional categorization research, our findings and approach have relevant implications for research in creative cognition. A central focus of the creative cognition approach has been to explain acts of creativity in terms of the mental representations and processes that are commonly studied in cognitive psychology and cognitive science \citep{finke1992creative,smith1995creative}. However, unlike other fields in the study of cognition, creative cognition research rarely employs quantitative models to evaluate the explanatory value of such representations and processes. Our modeling results provide a concrete example of how formal approaches may be used to gain insight into the nature of creative cognition.

In addition to demonstrating the utility of formal modeling in the study creative cognition, the PACKER model more specifically offers an additional interpretation of some of the field's most central findings. For example, perhaps the most foundational principle from this literature concerns the limiting influence of prior knowledge: individuals create new categories composed of features from existing classes, and what is created can be influenced drastically through the introduction of cues or examples \citep{marsh1999inadvertent,smith1993constraining}. In this paper, we have identified another important aspect of the constraining influence of prior knowledge: What is generated cannot be the same as what is already known. The results of our simulations with PACKER demonstrate that this influence is concisely explained in terms of a trade-off between within-category similarity and between-category dissimilarity.

Similarly, PACKER may offer an additional interpretation of existing accounts of creative generation. Most notably, a leading account within the creative cognition literature, the Path of Least Resistance \citep{ward1994structured,ward1995s}, also explains generation in terms of an exemplar-based retrieval process. This account was designed to explain the creative generation of natural categories (e.g., new species of plants and animals) and as a result relies strongly on the hierarchical organization of these categories: individuals are thought to retrieve an example of the higher-level category being generated (e.g., \textit{bird} may be retrieved from the category \textit{animal}), and then systematically alter what was retrieved to make something new. As the PACKER model does not assume knowledge is hierarchically organized \citep[this is true of the exemplar view more broadly, see][]{murphy2016exemplar}, the model may be viewed as a formal instantiation of the Path of Least Resistance for application in a traditional artificial categorization domain (in which there is no established hierarchy of categories). PACKER's success in explaining generation within an artificial domain motivates future work exploring the nature of category contrast within a more naturalistic setting.

@@

Finally, although an overall goal of this work is to advance the study of creative thought by explaining generation through the empirical and quantitative approaches from research in categorization, these approaches are not easily aligned with many of the myriad alternative theoretical viewpoints in the study of creativity \citep[for a review see][]{kozbelt2010theories}, such as those based on free association \citep{mednick1962associative} and conceptual combination \citep{estes2002emergence,murphy1988comprehending}. Instead, we reduced our focus by studying a highly complex behavior (creative generation) as it applies within a well-established domain (artificial category learning). Importantly, however, our findings do provide a clue about a more general, computational-level interpretation of creative generation. That is, in this paper we have made clear that contrast is a key constraint, as it is required in order to successful create something new. However, it cannot be the only constraint: It is not enough for a generated category to be distinct, as categories generated purely on the basis of contrast would not be distributionally coherent. Thus, a more general interpretation of creative generation would need to concern both the distinctness and coherence of what is created.

\nbcnote{I think importance sampling should fit in there pretty well.}
